Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-01-27 CAV 2025 TODO/experiments.tex
exp

%
We implemented \toolname\ in Python 3.8, and Gurobi 9.52 was used for solving LP and MILP problems. We conducted our evaluation on an AMD Threadripper 7970X  ($32$ cores$@4.0$GHz, 5nm) with 256 GB of main memory and 2 NVIDIA RTX 4090.

The  objectives of our evaluation was to answer the following  questions:

@@

\subsection{Evaluation of the Utility function to choose neurons to open}

%The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes,  to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.
To measure the impact of the utility function to select neurons to open, we focused on a small hard DNN, namely $5\times 100$, so as to be able to compute exact bounds for the first few layers using a full MILP encoding of the DNN for comparison purpose. We tested over the $\vx=59$th image in the MNIST dataset, as it has a large number of unstable ReLU nodes in the first few layers ($61$ in the first and $55$ in the second layer), so we can experiment with a larger choice of values) for $K=|X|$ the size of set $X$.
To measure the accuracy, we measure the uncertainty of all nodes in a layer:
the uncertainty of a node is the range between its computed lower and upper bound. 
We then average the uncertainty among all the nodes of the layer.
Formally, the uncertainty of a node $a$ with bounds $[\LB,\UB]$ is uncert$(a) = \UB - \LB$. The average uncertainty of layer $\ell$ is $\dfrac{\sum_{a\in l} \text{uncert}(a)}{size(\ell)}.$

@@

We focus on the uncertainty of nodes in the third layers, wrt ReLU nodes in the first and second layer. The bounds for nodes of the first two layers are computed exactly using the full MILP encoding. We report in Figure \ref{fig_table3} the average uncertainty of $\MILP_X$ following the choice of the $K$ heaviest neurons for our {\bf Utility} function, compared with a random choice, both for nodes exclusively from the previous Layer 2 or from both Layers 1 and 2. We compared with choosing based on strength$(n) = (\UB(n)-\LB(n))\cdot |W_{nz}|$ \cite{DivideAndSlide}, opening nodes in Layer 2 only.

@@

The Utility function selects very important neurons:
 to achieve the same accuracy than \cite{DivideAndSlide}, $2.5$ time fewer nodes (10 vs 25) are necessary when picking in the same previous Layer 2. The ability from {\bf Utility} to compare neurons from different layers enables even better frugality: $4$ time fewer nodes (5 vs 20, 10 vs 40) are necessary to reach the same accuracy than \cite{DivideAndSlide}.
Overall, choosing 35 neurons by {\bf Utility} improves accuracy by $95\%$ of what can be done if all $|X|=116$ nodes are opened compared with LP (corresponding to $|X|=0$).

\begin{figure}[h!]
	\centering
	\vspace*{-0.3cm}
	\includegraphics[height=8cm]{Plot for table 3}.
	\vspace*{-0.4cm}
	\caption{Average uncertainty of $\MILP_X$ for nodes of the third layer, for $X$ with $K$ ReLU nodes of the (1st and) 2nd layer, chosen by our {\bf Utility} function vs \cite{DivideAndSlide} vs random choice.}
	\label{fig_table3}
\end{figure}

>>>>

%
We implemented \toolname\ in Python 3.8, and Gurobi 9.52 was used for solving LP and MILP problems. We conducted our evaluation on an AMD Threadripper 7970X  ($32$ cores$@4.0$GHz) with 256 GB of main memory and 2 NVIDIA RTX 4090. The  objectives of our evaluation was to answer the following  questions:

@@

\subsection{Comparison between different scoring functions.}

%The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes,  to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.
To measure the impact of the scoring function to select neurons to open, 
we considered the largest complex CNN-B-Adv, with 20 000 nodes. 
We tested over the $\vx=37$th image in the CIFAR dataset.
To measure the accuracy, we measure the uncertainty of all nodes in the 4th layer:
the uncertainty of a node is the range between its computed lower and upper bound. 
We then average the uncertainty among all the nodes of the layer.
Formally, the uncertainty of a node $a$ with bounds $[\LB,\UB]$ is uncert$(a) = \UB - \LB$. The average uncertainty of layer $\ell$ is $\dfrac{\sum_{a\in l} \text{uncert}(a)}{size(\ell)}.$

\begin{figure}[b!]
	\centering
	\vspace*{-0.3cm}
	\includegraphics[height=7cm]{open.png}.
	\vspace*{-0.4cm}
	\caption{Average uncertainty of pMILP for neurons of the fourth layer of CNN-B-Adv, 
	when opening between 0 and 30 ReLUs, chosen randomly vs Huang \cite{DivideAndSlide} vs 
	$s_{FSB}$ vs $\Utility$ (with time-out=900s).}
	\label{fig_table3}
\end{figure}

@@

%We focus on the uncertainty of nodes in the third layers, wrt ReLU nodes in the first and second layer. The bounds for nodes of the first two layers are computed exactly using the full MILP encoding. 
We report in Figure \ref{fig_table3} the average uncertainty of $\MILP_X$ following the 
choice of the $K$ heaviest neurons of SAS, compared with a random choice, with Huang \cite{DivideAndSlide}, based on strength$(n) = (\UB(n)-\LB(n))\cdot |W_{nz}|$, and with GS.

@@

Overall, SAS is more accurate than GS, more accurate than Huang, more accurate than random choice of variables. SAS significantly outperforms other solutions, 
with around 3 times less binary variables for the same accuracy (10 v 30, 6 vs 16).
Each individual node of the last layer display a similar pattern, and the pattern is similar for different images (not displayed).

We also experimented on fixing an ordering on variable in Gurobi, with a fixed selection of nodes by SAS. Indeed, Gurobi relies on a BaB procedure to compute bounds. The standard Gurobi procedure is adaptative in each branch. We compared vs the fix order provided by SAS, as well as the fixed order provided by FSB, for both accuracy and runtime (lower is better).

We obtained the following:
Gurobi order:
30nodes : 1, 3.987387317631171, 260.4883162975311, 
45nodes : 1, 4.015747873604205, 265.53007888793945, 
60nodes : 1, 4.053423060042739, 262.1407461166382,

FSB order:
30nodes : 1, 3.9879578118285317, 246.69000577926636, 
45nodes : 1, 4.018561938508719, 247.65485453605652
60nodes : 1, 4.032634363809192, 247.81136560440063,

SAS order:
30nodes : 1, 3.9933637922653995, 264.4269104003906, 
45nodes : 1, 4.0307741092892035, 266.20811343193054, 
60nodes : 1, 4.060220618788057, 262.93414759635925,

Here,  SAS is as efficient as Gurobi, although the ordering is fixed, meaning that the ordering compensates for the non adaptivity. However, the fixed GS ordering (with the SAS selection) is even better, with slightly better accuracy and lower runtime. As explained earlier, this is due to the fact that global ordering is preferable when branching on choices far from the optimal solution.

### 2025-01-27 CAV 2025 TODO/formula.tex
exp

Assume that we want to compute an upper bound for neuron $z$ on layer $\ell_z$.
We write $n < z$ if neuron {\color{blue} $n$ is} on a layer before $\ell_z$, and $n \leq z$ if $n< z$ or $n=z$. We denote ($\Sol\_\max_X^z(n))_{n \leq z}$ a solution of $\mathcal{M}_X$ maximizing $z$. In particular, $\Sol\_\max_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider $(sol(n))_{n \leq z} = (\Sol\_\max_\emptyset^z(n))_{n \leq z}$, a solution maximizing the value for $z$ when all ReLU use the LP relaxation.
%(this can be obtained very efficiently by {\em one} call to an LP solver).
Function
$\Improve\_\max^z(n)=$ $\sol(z) - \Sol\_\max_{\{n\}}^z(z)$, 
accurately represents how much opening neuron $n < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_\max^z(n)\geq 0$ as $\Sol\_\max_{\{n\}}^z$ fulfills all the constraints of 
$\mathcal{M}_\emptyset$, so $\Sol\_\max_{\{n\}}^z(z) \leq \sol(z)$.
Similarly, we define ($\Sol\_\min_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_\min^z(n)$. Calling MILP on $\mathcal{M}_{\{n\}}$ for every neuron $n \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
{\color{blue} 
The main novelty of our SAS function is that it uses a (single) LP call to compute $(sol(n))_{n \leq z}$, with negligible runtime wrt the forthcoming  $\MILP_X$ call, and yet accurately approximates $\Improve\_\max^z(n)$ to choose a meaningful set $X$ of open nodes (Table \ref{tab:example1}).

@@

We can proceed inductively in the same way to define $\Utility\_\max^z(a)$ for deeper neurons $a$.

While $\Utility$ is a priori a more accurate way to select neurons to open in partial MILP (that will be verified in Table \ref{tab:example}), global scoring functions have two advantages: first, they are faster to compute (a unique backpropagation is sufficient, while SAS needs one full LP call to recover the solution, plus a forward propagation for each node $a$). This is not an issue for selecting nodes for partial MILP, as there is a single selection step. But this would be potentially more problematic for choosing branching nodes. A trade off could be to precompute $s_{FSB}$ and only on the most promising nodes run $\Utility$. 
The second shortcoming is the other side of the coin to consider improvement local to a solution: in the case of branching, one of the branch imposes to be far from the solution, 
while global scoring is as (in)accurate for both branches. We will verify that using GS for ordering nodes after the SAS selection is actually more efficient, for this very reason.

>>>>

Assume that we want to compute an upper bound for neuron $z$ on layer $\ell_z$.
We write $n < z$ if neuron $n$ is on a layer before $\ell_z$, and $n \leq z$ if $n< z$ or $n=z$. We denote ($\Sol\_\max_X^z(n))_{n \leq z}$ a solution of $\mathcal{M}_X$ maximizing $z$. In particular, $\Sol\_\max_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider $(sol(n))_{n \leq z} = (\Sol\_\max_\emptyset^z(n))_{n \leq z}$, a solution maximizing the value for $z$ when all ReLU use the LP relaxation.
%(this can be obtained very efficiently by {\em one} call to an LP solver).
Function
$\Improve\_\max^z(n)=$ $\sol(z) - \Sol\_\max_{\{n\}}^z(z)$, 
accurately represents how much opening neuron $n < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_\max^z(n)\geq 0$ as $\Sol\_\max_{\{n\}}^z$ fulfills all the constraints of 
$\mathcal{M}_\emptyset$, so $\Sol\_\max_{\{n\}}^z(z) \leq \sol(z)$.
Similarly, we define ($\Sol\_\min_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_\min^z(n)$. Calling MILP on $\mathcal{M}_{\{n\}}$ for every neuron $n \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
{\color{blue} 
The main novelty of our SAS function is that it uses a (single) LP call to compute $(sol(n))_{n \leq z}$, with negligible runtime wrt the forthcoming  $\MILP_X$ call, and yet accurately approximates $\Improve\_\max^z(n)$ to choose a meaningful set $X$ of open nodes (Fig. \ref{fig_table3}).

@@

We proceed inductively to define $\Utility\_\max^z(a)$ for deeper neurons $a$.

While $\Utility$ is a priori a more accurate way to select neurons to open in partial MILP (that will be verified in Fig. \ref{fig_table3}), global scoring functions have two advantages: first, they are faster to compute (a unique backpropagation is sufficient, while SAS needs one full LP call to recover the solution, plus a forward propagation for each node $a$). This is not an issue for selecting nodes for partial MILP, as there is a single selection step. This would be potentially problematic for choosing branching nodes in BaB, at each branch. A trade off could be to precompute FSB and only on the most promising nodes run $\Utility$. The second shortcoming is the other side of the coin to consider improvement local to a solution: in the case of branching, one of the branch imposes to be far from the solution and $\Utility$ is not accurate on that branch, while global scoring is as (in)accurate for both branches. We will verify that using GS for ordering nodes after the SAS selection is actually more efficient, for this very reason.
